Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-23 article.tex
Damien & Arnaud thoughts and writing

Let us assume that the fluorescence concentrations of all $p$ neurons modeled
as a set of random variables $X = \{X_1, \dots, X_p\}$ are drawn from a time
invariant joint probability distribution $P_X$.
In the probability framework, two random variables $X_i$ and $X_j$ are
said to be conditionally independent to a (possibly empty) set of controlling
random variables $Z$, denoted $X_i \perp X_j | Z$, if the joint conditional
probability distribution $P_{X_i, X_j | Z}$ factorizes into $P_{X_i|Z}
P_{X_j|Z}$. In particular, when $X_i$ and $X_j$ are conditionally independent
with respect to all the other variables, denoted $X_i \perp X_j | X^{-i,j}$,
then the observation of $X_i$ given the knowledge of $X^{-i,j}$ provides no
information about $X_j$, and vice-versa. From a graphical point of view and
assuming no hidden variable, this indicates that $X_i$ and $X_j$ do not
directly interact (i.e., $(i,j) \notin E$). By opposition, if $X_i$ and $X_j$
are conditionally dependent with respect to all other variables, denoted $X_i
\not\perp X_j | X^{-i,j}$, then the observation of one of the two variables
provides some information about the other, therefore highlighting a direct
interaction between the two random variables (i.e., $(i, j) \in E$).

Under the assumption that random variables follow a joint Gaussian
distribution $\mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and covariance
matrix $\Sigma$, $X_i$ and $X_j$ are conditionally independent with respect to
the other variables $X^{-i,j}$ if their \textit{partial correlation} statistic
is null~\citep{baba2004partial}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j |
X^{-i,j}} \neq 0$, where $\rho_{X_i, X_j | X^{-i,j}} \in [-1;1]$ can now be
interpreted as the degree of association between the two variables (i.e., the larger the
value, the stronger the interaction). In practice due to the finite size of
time series, noise and modeling assumptions, the partial correlation
$\rho_{X_i, X_j | X^{-i,j}}$ will never be zero. If instead of a degree
of association, one would want to infer the underlying graph a statistical
test would be required to assert conditional independences. Note also that
the partial correlation
statistic is symmetric (i.e., $\rho_{X_i, X_j | X^{-i,j}} = \rho_{X_j, X_i |
X^{-j,i}}$) and therefore does not account for the direction of the
interaction. Together, the partial correlation statistics correspond to the
inverse $\Sigma^{-1}$ of the covariance matrix, also known as the concentration
or precision matrix. Computationally, they can be estimated efficiently in
$O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal components
extracted through PCA~\citep{bishop2006pattern} on the processed data. In our
experiments, $M$ is set to $0.8 p$, thereby allowing us to further filter the
the data by removing the contribution of principal components
likely due to noise.

>>>>

Our  connection inference  procedure  relies  on several  assumptions. First,
we will assume that  the fluorescence concentrations of all $p$ neurons can be
modeled as a set of random variables $X = \{X_1, \dots, X_p\}$  that are drawn
from a time invariant joint  probability distribution $P_X$. Or, say otherwise,
we assume that there exists a probability distribution $P_X$ such that the
fluorescence concentrations at a particular time $t$ can be seen as an
independent realization of $P_X$.

To a joint probability distribution $P_X$ corresponds a minimal dependency
graph. Each node of this graph is a random variable. Two random variables are
only connected if they are dependent. We consider that two neurons are
connected if  and only if,  there exists  in this dependency  graph an edge
that connects  directly the  two random  variables to  which they correspond.
This is equivalent to writing that two neurons $i$ and $j$ are  not  directly
connected  if and only if  $X_i$  and $X_j$  are  conditionally independent
with  respect to all  other variables. We remind  that two random  variables
$X_i$  and  $X_j$  are  said  to  be  conditionally independent to a (possibly
empty)  set of controlling random variables $Z$, denoted $X_i \perp X_j | Z$,
if the joint conditional probability distribution   $P_{X_i,  X_j   |   Z}$
factorizes  into   $P_{X_i|Z} P_{X_j|Z}$.  In particular,  when  $X_i$ and
$X_j$ are  conditionally independent  with respect  to all  the other
variables, denoted  $X_i \perp  X_j  | X^{-i,j}$,  then  the  observation  of
$X_i$  given  the knowledge  of  $X^{-i,j}$ provides  no  information  about
$X_j$,  and vice-versa.

In the  particular case   where $P_X$ is  a joint  Gaussian distribution
$\mathcal{N}(\mu; \Sigma)$ of mean  vector $\mu$ and covariance matrix
$\Sigma$, $X_i$  and $X_j$ are conditionally  independent with respect to   the
other   variables   $X^{-i,j}$   if  their   \textit{partial   correlation}
statistic is null~\citep{baba2004partial}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j |
X^{-i,j}} \neq 0$, where $\rho_{X_i, X_j | X^{-i,j}} \in [-1;1]$. Note that the
larger the value of this term, the stronger the dependence between variables
$X_i$ and $X_j$.

Our approach implicitly makes the additional assumption  that $P_x$ is  a joint
Gaussian distribution since it  outputs as  degree of association between two
neurons $i$ and $j$ an estimate of $\rho_{X_i, X_j | X^{-i,j}}$. Note also that
the partial correlation statistic is symmetric  (i.e., $\rho_{X_i, X_j |
X^{-i,j}} = \rho_{X_j, X_i | X^{-j,i}}$) and therefore does not account for the
direction of the interaction.

We detail now our procedure for estimating $\rho_{X_i, X_j | X^{-i,j}}$ from
the data. The partial correlation statistics is computed through the inverse
$\Sigma^{-1}$ of the covariance matrix $\Sigma$, also known as the
concentration or precision matrix. Computationally, they can be estimated
efficiently in $O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal
components extracted through PCA~\citep{bishop2006pattern} on the processed
data. In our experiments, $M$ is set to $0.8 p$, thereby allowing us to further
filter the the data by removing the contribution of principal components likely
due to noise.
